Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2019-11-28 paper.Rnw
john's sec 2 comments

Since our goal was to better understand the CTA1
treatment effect, we focused our analysis on data from high school students in
the second year of the CTA1 trial, for whom the treatment effect was
most evident.

We merged data from two sources: covariate, treatment and outcome data gathered by RAND, and computerized log data gathered by Carnegie Learning.
Table \ref{tab:covariateBalance} describes the covariates we used, including
missingness information, control and treatment means, and standardized
differences \citep[c.f.][]{kalton1968standardization}.
We singly-imputed missing values with the Random Forest routine implemented
by the \texttt{missForest} package in \texttt{R}
\citep{missForest,rcite},  which estimated the ``out of box'' imputation
errors also shown in Table \ref{tab:covariateBalance} as part of the random forest regression.

@@

Over the course of the effectiveness study, Carnegie Learning gathered
log data for each worked problem.
The resulting dataset lists the problem name, section, and unit of
each problem, the numbers of hints requested and errors committed, and time-stamps.
Log data were missing for some students, either because the log files were not retrievable, or because of an imperfect ability to link log data to other student records.
Treatment schools with mastery data missing
for 90\% or more students were omitted from the analysis, along with
their entire randomization block.
Of the remaining \Sexpr{Nt} students, \Sexpr{round(obsUse/Nt*100)}\% (\Sexpr{prettyNum(obsUse,',')}) had log
data.
For the sake of simplicity, the treatment students without log data
were dropped from the study.\footnote{Including these students in a
  principal stratification model is straightforward \citep{aoas}; see
  the online supplement for results from a model including them, which
  were broadly similar to the results presented here. Including
  subjects with missing log data in a mediation or observational study
  design can be more problematic \citep[see, e.g.][]{li2017identifiability}.}
Problem log data for sections that were not part of the
standard CTA1 Algebra I curriculum and problems worked by fewer than 100
students were omitted from the
dataset.

>>>>

\citep{pane2014effectiveness} estimated effects separately in middle
schools and high schools in each the first two years of the RCT; the
only statistically significant treatment effect was in the high school
sample in the second year.
Since our goal was to better understand the CTA1
treatment effect, we focused our analysis on data from high school students in
the second year of the CTA1 trial, for whom the treatment effect was
most evident.

We merged data from two sources: computerized log data gathered by
Carnegie Learning, and covariate, treatment and outcome data gathered by RAND.

The log data lists the problem name, section, and unit of
each problem, the numbers of hints requested and errors committed, and time-stamps.
Log data were missing for some students, either because the log files
were not retrievable, or because of an imperfect ability to link log
data to other student records.
Further, log data for sections that were not part of the
standard CTA1 Algebra I curriculum and problems worked by fewer than 100
students were omitted from the
dataset.

To construct our analysis dataset, we first dropped treatment schools with log data missing
for 90\% or more students.
Prior to randomization, schools were stratified into pairs or triples
based on baseline covariates, and randomized between treatment and
control condition within these randomization blocks.
When we dropped a treatment school from our analysis, we also dropped
the control schools in its randomization block.
Of the remaining \Sexpr{prettyNum(Nt,',')} students, \Sexpr{round(obsUse/Nt*100)}\% (\Sexpr{prettyNum(obsUse,',')}) had log
data.
Then, for the sake of simplicity, the treatment students without log data
were dropped from the study.\footnote{Including these students in a
  principal stratification model is straightforward \citep{aoas}; see
  the online supplement for results from a model including them, which
  were broadly similar to the results presented here. Including
  subjects with missing log data in a mediation or observational study
  design can be more problematic \citep[see, e.g.][]{li2017identifiability}.}

Table \ref{tab:covariateBalance} describes the covariates we used, including
missingness information, control and treatment means, and standardized
differences \citep[c.f.][]{kalton1968standardization} from the final
analysis sample.
We singly-imputed missing values with the Random Forest routine implemented
by the \texttt{missForest} package in \texttt{R}
\citep{missForest,rcite},  which estimated ``out of box'' imputation
error rates as part of the random forest regression, also shown in Table \ref{tab:covariateBalance}.

@@
